---
title: "Alpha and Beta Diversity Analysis"
author: "Yixuan Yang"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    self_contained: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Set Up

```{r message=FALSE}
rm(list = ls())
set.seed(2024)
library(tidyverse)
library(here)
library(phyloseq)
library(vegan)
library(rstatix)
theme_set(theme_bw())
options(digits=4)
max.core <- parallel::detectCores()

ps.rare <- readRDS(here('data','following_study','ps_rarefied.rds')) 
sample_data(ps.rare)$Shannon <- estimate_richness(ps.rare)$Shannon
# transform data into proportion
ps.rare.prop <- ps.rare %>% transform_sample_counts(function(x) x/sum(x))

sam <- data.frame(sample_data(ps.rare))
```

# Functions

`plot_ord` draws ordination plot for different factors using `plot_ordination` function in `phyloseq` package.

`permanova` performs permutational multivariate analysis of variance (PERMANOVA) based on `adonis2` function in `vegan` package.

`permdisp` performs permutational analysis of multivariate dispersions (PERMDISP) based on `betadisper` function in `vegan` package.

```{r}
plot_ord <- function(data, factor, method, distance){
    data.ord <- ordinate(data, method = method, distance = distance)
    p <- plot_ordination(data, data.ord, color = factor)
    p <- p + stat_ellipse(type = "t",geom = "polygon",alpha = 0)
    p <- p + ggtitle(str_c(factor,method,distance, sep = ' - '))
    print(p)
}

permanova <- function(data, formula, method, permutations=1e4, strata = NULL, core = max.core){
    message('PERMANOVA Model: ', method, '~', formula, '; Strata: ', ifelse(is_null(strata), 'None', as.character(strata)))
    dist.matrix <- phyloseq::distance(data, method=method)
    df <- data.frame(sample_data(data))
    model <- as.formula(paste0('dist.matrix~', formula))
    if (!is_null(strata)) {strata <- df[,strata]}
    result <- adonis2(model,
                      data = df,
                      permutations=permutations,
                      strata = strata,
                      parallel = core,
                      by = 'term',
                      na.action = na.omit)
    return(result)
}
permdisp <- function(data, group, method, permutations=1e4, pairwise = FALSE, core = max.core){
    message('PERMDISP Model: ', method, '~', group)
    dist.matrix <- phyloseq::distance(data, method=method)
    df <- data.frame(sample_data(data))
    beta.disp <- betadisper(dist.matrix, group = df[,group])
    result <- permutest(beta.disp, permutations = permutations, pairwise = pairwise, type = 'centroid')
    return(result)
}
```

# Start

In this section, we want to estimate the effect of different factors on the microbial diversity. The factors we are focusing on are `Household`, `Epileptic.or.Control`, `Breed.Group..1.`, `Pheno.Y.N`, `Sex`, and `Age..months.`. We compare the species richness (Shannon index) among different factors using ANOVA, and compare the centroid of dissimilarity of microbial community between different groups using PERMANOVA using the Bray-Curtis and weighted Unifrac distance, and visualized using multi-dimensional scaling. PERMDISP was used to test the homogeneity of multivariate dispersions among groups.

# Household Effect

## Alpha Diversity

```{r}
ggplot(sam,aes(x = as.numeric(Household), y = Shannon, group = Household)) +
    geom_point() + geom_line() + xlab('Household')
anova_test(Shannon~Household, data = sam, type = 1)
```

Here we see the Shannon diversity index is significantly different among households.

## Beta Diversity

### Bray-Curtis distance

```{r}
permanova(ps.rare.prop, 'Household', 'bray')
```

### Weighted-Unifrac distance

```{r}
permanova(ps.rare.prop, 'Household', 'wunifrac')
```

# Epileptic Effect

## Alpha Diversity

```{r}
ggplot(sam, aes(x = Epileptic.or.Control, y = Shannon)) + 
    geom_boxplot() + geom_jitter(height = 0, width = 0.25)
anova_test(Shannon~Household+Epileptic.or.Control, data = sam, type = 1)
```

## Beta Diversity

### Bray-Curtis distance

```{r}
plot_ord(ps.rare.prop, 'Epileptic.or.Control','MDS','bray')
plot_ord(ps.rare.prop, 'Epileptic.or.Control','NMDS','bray')
permanova(ps.rare.prop, 'Epileptic.or.Control', 'bray', strata = 'Household')
permdisp(ps.rare.prop, 'Epileptic.or.Control', 'bray')
```

### Weighted-Unifrac distance

```{r}
plot_ord(ps.rare.prop, 'Epileptic.or.Control','MDS','wunifrac')
plot_ord(ps.rare.prop, 'Epileptic.or.Control','NMDS','wunifrac')
permanova(ps.rare.prop, 'Epileptic.or.Control', 'wunifrac', strata = 'Household')
permdisp(ps.rare.prop, 'Epileptic.or.Control', 'wunifrac')
```


# Breed Effect

## Alpha Diversity

```{r}
sam.breed <- sam %>%
  filter(is.na(Breed.Group..1.) == FALSE) %>% 
  filter(is.na(Breed.Group..2.) == TRUE, is.na(Breed.Group..3.) == TRUE)
nrow(sam.breed)
  
ggplot(sam.breed) +
    geom_point(aes(x = Breed.Group..1., y = Shannon, colour = Breed.Group..1.)) +
    facet_wrap(~Epileptic.or.Control) + 
    theme(axis.text.x = element_blank(), axis.ticks.x.bottom = element_blank())

anova_test(Shannon~Household + Breed.Group..1., data = sam.breed, type = 1)
```

## Beta Diversity

### Bray-Curtis distance

```{r}
plot_ord(ps.rare.prop, 'Breed.Group..1.','MDS','bray')
plot_ord(ps.rare.prop, 'Breed.Group..1.','NMDS','bray')
permanova(ps.rare.prop, 'Household + Breed.Group..1.', 'bray')
permdisp(ps.rare.prop, 'Breed.Group..1.', 'bray')
```

### Weighted-Unifrac distance

```{r}
plot_ord(ps.rare.prop, 'Breed.Group..1.','MDS','wunifrac')
plot_ord(ps.rare.prop, 'Breed.Group..1.','NMDS','wunifrac')
permanova(ps.rare.prop, 'Household + Breed.Group..1.', 'wunifrac')
permdisp(ps.rare.prop, 'Breed.Group..1.', 'wunifrac')
```

# Drug Effect

## Alpha Diversity

```{r}
sam.drug <- sam %>% filter(Epileptic.or.Control == 'Epileptic')
ggplot(sam.drug, aes(x = Pheno.Y.N, y = Shannon)) + 
    geom_boxplot() + geom_jitter(height = 0, width = 0.25)
anova_test(Shannon~Pheno.Y.N, data = sam.drug, type = 1)
```

## Beta Diversity

### Bray-Curtis distance

```{r}
ps.drug <- ps.rare.prop %>% subset_samples(Epileptic.or.Control == 'Epileptic')
plot_ord(ps.drug, 'Pheno.Y.N','MDS','bray')
plot_ord(ps.drug, 'Pheno.Y.N','NMDS','bray')
permanova(ps.drug, 'Pheno.Y.N', 'bray')
permdisp(ps.drug, 'Pheno.Y.N', 'bray')
```

### Weighted-Unifrac distance

```{r}
plot_ord(ps.drug, 'Pheno.Y.N','MDS','wunifrac')
plot_ord(ps.drug, 'Pheno.Y.N','NMDS','wunifrac')
permanova(ps.drug, 'Pheno.Y.N', 'wunifrac')
permdisp(ps.drug, 'Pheno.Y.N', 'wunifrac')
```


# Sex Effect

```{r}
prop.test(xtabs(~Household+Sex, data = sam))
```

## Alpha Diversity

```{r}
ggplot(sam, aes(x = Sex, y = Shannon)) + 
    geom_boxplot() + geom_jitter(height = 0, width = 0.25)
anova_test(Shannon~Household+Sex, data = sam, type = 1)
```

## Beta Diversity

### Bray-Curtis distance

```{r}
plot_ord(ps.rare.prop, 'Sex','MDS','bray')
plot_ord(ps.rare.prop, 'Sex','NMDS','bray')
permanova(ps.rare.prop, 'Household+Sex', 'bray')
permdisp(ps.rare.prop, 'Sex', 'bray')
```

### Weighted-Unifrac distance

```{r}
plot_ord(ps.rare.prop, 'Sex','MDS','wunifrac')
plot_ord(ps.rare.prop, 'Sex','NMDS','wunifrac')
permanova(ps.rare.prop, 'Household+Sex', 'wunifrac')
permdisp(ps.rare.prop, 'Sex', 'wunifrac')
```

# Age Effect

## Alpha Diversity

```{r}
sam[which(is.na(sam$Age..months.)),'Household'] 
# remove households that have dog with unspecific age
sam.Age <- sam %>% filter(!(Household %in% c('9', '24')))
ggplot(sam.Age,) +
    geom_line(aes(x = as.numeric(Household), y = Age..months., group = Household)) + 
    geom_point(aes(x = as.numeric(Household), y = Age..months., group = Household, colour = Epileptic.or.Control)) +
    xlab('Household') + ylab('Age in month')

anova_test(Shannon~Household+Age..months., data = sam.Age, type = 1)
```

## Beta Diversity

### Bray-Curtis distance

```{r}
ps.age <- ps.rare.prop %>% subset_samples(!(Household %in% c('9', '24')))
permanova(ps.age, 'Age..months.', 'bray', strata = 'Household')
```

### Weighted-Unifrac distance

```{r}
permanova(ps.age, 'Age..months.', 'wunifrac')
```

```{r}
sessioninfo::session_info()
```