---
title: "Analyze Collinsella"
author: "Yixuan Yang"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    self_contained: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Set Up

```{r message=FALSE}
set.seed(2024)
library(tidyverse)
library(here)
library(rstatix)
library(phyloseq)
theme_set(theme_bw())

ps.rare <- readRDS(here('data','following_study','ps_rarefied.rds')) %>%
    filter_taxa(function(x) {
        prevlance <- sum(x > 0) / length(x)
        prevlance > 0.1
        }, prune = TRUE)
```

# Start

In this section, we are mainly focusing on 

We have found that `Collinsella` is the top candidate that influence dog's microbiome. Here we investigate its existence.

```{r}
ps.collinsella <- ps.rare %>%
    subset_taxa(Genus == 'Collinsella') 

# add genus level collinsella count to otu table
otu.df <- otu_table(ps.collinsella) %>% 
    cbind(Genus = rowSums(otu_table(ps.collinsella))) %>%
    cbind(data.frame(sample_data(ps.collinsella))) %>% 
    pivot_longer(c(starts_with('ASV'), 'Genus'), names_to = 'ASV', values_to = 'Abundance')


lfc <- otu.df %>%
    group_by(Household, ASV) %>%
    summarise(LFC = log2(Abundance[Epileptic.or.Control == "Epileptic"] + 1) -
                  log2(Abundance[Epileptic.or.Control == "Control"] + 1),
              .groups = 'drop')

ggplot(lfc) +
    geom_bar(aes(x = as.numeric(Household), y = LFC, fill = (LFC > 0)), stat = 'identity') +
    facet_wrap(~ASV, scales = 'free_y') +
    labs(x = 'Household', y = 'log 2 fold change', fill = 'LFC > 0')
```

Are there always more Collinsella ASVs in Epileptic group than control group?

```{r}
otu.df %>% 
    arrange(Household) %>%
    group_by(ASV) %>%
    pairwise_sign_test(Abundance ~ Epileptic.or.Control, ref.group = 'Epileptic', alternative = 'greater')
```


# Contingency Tables

Existence frequency of Collinsella across Epi v.s. Control groups

```{r}
ftable(xtabs(sign(Abundance) ~ Epileptic.or.Control + ASV, data = otu.df))
```

Epileptic cross sex table for each Collinsella ASVs

```{r}
ftable(xtabs(sign(Abundance) ~ Sex  + Epileptic.or.Control + ASV, data = otu.df))
```

Does Epileptic or Control group associate with sex for Collinsella ASV?

```{r}
otu.df %>% 
    split(.$ASV) %>% 
    lapply(function(df) summary(xtabs(sign(Abundance) ~ Sex  + Epileptic.or.Control, data = df)))
```

```{r}
ftable(xtabs(sign(Abundance) ~ Sex  + Pheno.Y.N + ASV, data = otu.df))
```

Does Drug usage associate on sex to the existence frequency of Collinsella?

```{r}
otu.df %>% 
    filter(Epileptic.or.Control == 'Epileptic') %>% 
    ggplot() +
    geom_boxplot(aes(x = Pheno.Y.N, y = Abundance)) +
    facet_wrap(~ASV, scale = 'free_y')


otu.df %>% 
    filter(Epileptic.or.Control == 'Epileptic') %>% 
    split(.$ASV) %>% 
    lapply(function(df) chisq.test(xtabs(sign(Abundance) ~ Pheno.Y.N, data = df)))
```

# Negative Binomial Regression Model

```{r}
otu.df %>%
    split(.$ASV) %>%
    lapply(function(df) MASS::glm.nb((Abundance + 1) ~ Age..months., data = df) %>% summary)
```

