---
title: "Differential Abundance Analysis"
author: "Yixuan Yang"
output:
  html_document:
    toc: true
    toc_float: true
    self_contained: true
---

# Set Up

```{r, message=FALSE}
rm(list = ls())
set.seed(2024)
library(tidyverse)
library(phyloseq)
library(here)
library(gt)
library(GGally)
library(ggcorrplot)
theme_set(theme_bw())

# DA tools
library(ALDEx2)
library(ANCOMBC)
library(corncob)
library(MicrobiomeStat) # for Linda

ps <- readRDS(here('data','following_study','ps.rds'))
ps.rare <- readRDS(here('data','following_study','ps_rarefied.rds'))
ps.present <- ps %>% transform_sample_counts(sign)

alpha <- 0.05
max.cores <- parallel::detectCores()
```

# Taxa Filtering

Before the DA analysis, we want to filter the data by prevlence and minimum read depth. Here we are showing a tracing plot of the remained read and taxa for different prevlence threshold with minimum read depth of 50.

```{r} 
# test for filtering threshold
test.filter <- function(ps, prevlence, min_read = 50) {
    sub.ps <- ps %>%
      filter_taxa(function(x) sum(x)>min_read & sum(x>0)/length(x) >= prevlence, prune = TRUE)
    retained.read <- otu_table(sub.ps) %>% rowSums() %>% mean()
    retained.taxa <- otu_table(sub.ps) %>% ncol()
    orig.read <- otu_table(ps) %>% rowSums() %>% mean()
    orig.taxa <- otu_table(ps) %>% ncol()
    return(data.frame(taxon_size = retained.taxa/orig.taxa, average.read.depth = retained.read/orig.read))
}
trace <- map_dfr(seq(0,0.9, 0.1), function(i) test.filter(ps=ps, prevlence=i))

ggplot(trace, aes(x = average.read.depth, y = taxon_size)) +
    geom_point() +
    geom_line() + 
    scale_x_reverse(labels = scales::percent) +
    scale_y_continuous(labels = scales::percent) +
    xlab('percentage of reads remained') +
    ylab('percentage of ASV remained') 
```

By setting the minimum prevalence to 0.1 and minimum read depth of 50, we kept 13% ASVs with 96% reads remained.

```{r} 
head(trace)
keep <- ps %>% 
  filter_taxa(function(x) sum(x)>100 & sum(x>0)/length(x)>0.1, prune = TRUE) %>%
  tax_table() %>% rownames()
```

Basic information of non-rarefied data

```{r} 
ps <- ps %>% prune_taxa(keep, .)
ps
```

Basic information of rarefied data

```{r} 
ps.rare <- ps.rare %>% prune_taxa(keep, .)
ps.rare
```

# Differential Abundance Analysis

## ALDEx2

### pairwise T-test

```{r message=FALSE}
clr.data <- aldex.clr(t(otu_table(ps)), as.character(sample_data(ps)$Epileptic.or.Control), mc.samples=128, denom="all")

aldex.t.test.p <- aldex.ttest(clr.data, paired.test = TRUE)
aldex.t.test.effect <- aldex.effect(clr.data, paired.test = TRUE, useMC = TRUE, verbose = FALSE)

if (!identical(rownames(aldex.t.test.p), rownames(aldex.t.test.effect))) stop('taxa names are not matched')
aldex.t.test <- data.frame(aldex.t.test.p, aldex.t.test.effect) %>% 
  rownames_to_column('taxon') %>%
  mutate(LFC = rab.win.Epileptic - rab.win.Control) %>% 
  mutate(p.value = we.ep) %>%
  dplyr::select(taxon, LFC, p.value)
    
aldex.t.test %>% arrange(str_rank(taxon, numeric = TRUE)) %>% head()
```

### GLM test

```{r message=FALSE}
model <- model.matrix(~ Epileptic.or.Control + Household, data.frame(sample_data(ps)))
clr.data <- aldex.clr(t(otu_table(ps)), model, mc.samples=128, denom="all")

aldex.glm.test.p <- aldex.glm(clr.data, model)
aldex.glm.test.effect <- aldex.glm.effect(clr.data, verbose = FALSE, useMC = TRUE)$Epileptic.or.ControlEpileptic
if (!identical(rownames(aldex.glm.test.p), rownames(aldex.glm.test.effect))) stop('taxa names are not matched')

aldex.glm.test <- data.frame(p.value = aldex.glm.test.p$`Epileptic.or.ControlEpileptic:pval`, aldex.glm.test.effect) %>% 
    rownames_to_column('taxon') %>%
    mutate(LFC = rab.win.1 - rab.win.0) %>% # 0 is control, 1 is epileptic, set is `data-cleaning.Rmd`
    dplyr::select(taxon, LFC, p.value)

aldex.glm.test %>% arrange(str_rank(taxon, numeric = TRUE)) %>% head()
```

## ANCOMBC

```{r message=FALSE}
ancombc.test <- ancombc(data = ps,
                        formula = 'Epileptic.or.Control + Household',
                        taxa_are_rows = FALSE,
                        prv_cut = 0,
                        conserve = TRUE,
                        n_cl = max.cores)

ancombc.test <- data.frame(taxon = ancombc.test$res$lfc$taxon,
                           # convert logfold change to log2fold change
                           LFC = log2(exp(ancombc.test$res$lfc$Epileptic.or.ControlEpileptic)),
                           p.value = ancombc.test$res$p_val$Epileptic.or.ControlEpileptic)

ancombc.test %>% arrange(str_rank(taxon, numeric = TRUE)) %>% head
```

## corncob

```{r message=FALSE}
corncob.test <- differentialTest(formula= ~ Epileptic.or.Control + Household,
                                 phi.formula = ~ 1,
                                 formula_null = ~ Household,
                                 phi.formula_null = ~ 1,
                                 test = 'LRT',
                                 fdr_cutoff = 1, # don't filter by FDR
                                 data = ps,
                                 taxa_are_rows = FALSE)

# convert corcob result to log2fold difference
corncob.extract.result <- function(corncob.da, i) {
    regression.result <- corncob.da$all_models[[i]]$coefficients
    # p value
    p <- corncob.da$p[i]
    # relative abundance
    RA.ctl <- regression.result['mu.(Intercept)','Estimate'] %>% invlogit()
    RA.epi <- regression.result[c('mu.(Intercept)','mu.Epileptic.or.ControlEpileptic'),'Estimate'] %>% sum() %>% invlogit()
    LFC <- RA.epi - RA.ctl # NOTE: this is not log2fold change, but the difference of relative abundance, since corncob does not provide lfc.
    summary <- data.frame(LFC = LFC, p.value = p) %>% rownames_to_column('taxon')
    return(summary)
}
corncob.test <- map_dfr(1:length(taxa_names(ps)), function(i) corncob.extract.result(corncob.test, i)) 
corncob.test %>% 
  arrange(str_rank(taxon, numeric = TRUE)) %>%
  mutate(`relative abundance difference`= LFC) %>% dplyr::select(-LFC) %>% head
```

## LinDA

```{r}
linda.test <- linda(t(data.frame(otu_table(ps))),
                   data.frame(sample_data(ps)),
                   formula = '~ Epileptic.or.Control + Household',
                   feature.dat.type = 'count',
                   n.cores = max.cores,
                   verbose = FALSE)

linda.test <- linda.test$output$Epileptic.or.ControlEpileptic %>% 
    dplyr::select(log2FoldChange, pvalue) %>%
    rownames_to_column('taxon') 
colnames(linda.test)[2:3] <- c('LFC', 'p.value')

linda.test %>% arrange(str_rank(taxon, numeric = TRUE)) %>% head()
```

## pairwise T-test on rarefied data

```{r}
paired.t.test <- function(data) {
    splited.data <- data %>% 
        arrange(Household, Epileptic.or.Control) %>% 
        split(.$Epileptic.or.Control)
    t.test.result <- t.test(splited.data[['Epileptic']]$Abundance,
                            splited.data[['Control']]$Abundance,
                            paired = TRUE, alternative = 'two.sided')
    data.frame(taxon = unique(data$OTU),
               LFC = log2(mean(splited.data[['Epileptic']]$Abundance + 0.5)) - log2(mean(splited.data[['Control']]$Abundance + 0.5)),
               p.value = t.test.result$p.value)
}

pairwise.t.test <- psmelt(ps.rare) %>%
    group_by(OTU) %>%
    group_split() %>%
    map_dfr(paired.t.test)

pairwise.t.test %>% arrange(str_rank(taxon, numeric = TRUE)) %>% head()
```


# Comparative Analysis

Here we compare the results of different methods. We first show the top 10 ASVs with the smallest p-value for each method.

```{r}
test.results <- list("ALDEx2 t test" = aldex.t.test,
                     "ALDEx2 GLM" = aldex.glm.test,
                     "ANCOM-BC" = ancombc.test,
                     "corncob" = corncob.test,
                     "LinDA" = linda.test, 
                     "paired t test" = pairwise.t.test) %>% 
    bind_rows(.id = 'method') %>% 
    group_by(method) %>% arrange(str_rank(taxon, numeric = TRUE)) %>% ungroup()

tax <- tax_table(ps) %>% 
    as.data.frame() %>% 
    dplyr::select(Phylum, Genus) %>% 
    rownames_to_column('taxon')

p.value.tb <- test.results %>% 
    dplyr::select(taxon, method, p.value) %>% 
    pivot_wider(names_from = method, values_from = p.value) %>%
    rowwise() %>%
    mutate(SignificantCount = sum(c_across(-taxon) < alpha))

top.p <- p.value.tb %>%
    left_join(tax,., by = 'taxon') %>% 
    group_by(SignificantCount) %>% 
    arrange(str_rank(taxon, numeric = TRUE), .by_group = TRUE) %>%
    arrange(desc(SignificantCount)) %>%
    ungroup() %>% 
    dplyr::select(-SignificantCount)

tb <- top.p %>% 
    head(10) %>% 
    gt() %>% 
    fmt_number(decimals = 4) %>% 
    data_color(columns = colnames(top.p)[4:9],
               colors = scales::col_bin(palette = c(scales::hue_pal()(1), "transparent"),
                                        bins = c(-Inf, 0.05, Inf)))
tb
writexl::write_xlsx(top.p, here('data', 'following_study', 'top_ASV.xlsx'))
```

Does the effect size of each method agreed in direction? The plot below shows that including `corncob` all methods have consistent direction of effect size.

```{r}
test.results$taxon <- factor(test.results$taxon,
                             levels = top.p$taxon)
test.results %>%
    mutate(direction = if_else(LFC > 0, 'positive', 'negative')) %>% 
    filter(taxon %in% head(top.p$taxon, 10)) %>%
    ggplot() +
    geom_tile(aes(x = method, y = taxon, fill = direction), color = 'black') +
    theme(axis.text.x = element_text(angle = 60, hjust = 1))
```


Effect size of top 10 ASVs for each method.

```{r}
effect.plot <- test.results %>% 
    filter(method != 'corncob') %>% 
    filter(taxon %in% head(top.p$taxon, 10)) %>% 
    ggplot() +
        geom_bar(aes(x = taxon, y = LFC, fill = method),
                 position="dodge", stat = 'identity', width = 0.7) +
        theme(axis.text.x = element_text(angle = 60, hjust = 1)) 
effect.plot
```

correlation of p-value for each method.

```{r}
top.p %>%  
    dplyr::select(-c(Phylum, Genus)) %>%
    column_to_rownames('taxon') %>% 
    ggpairs(progress = FALSE)

cor.plot <- top.p %>%  
    dplyr::select(-c(Phylum, Genus)) %>%
    column_to_rownames('taxon') %>%
    cor() %>%
    ggcorrplot(lab = TRUE, type = 'lower', show.legend = FALSE, tl.cex = 10)
cor.plot
```

```{r include=FALSE}
library(patchwork)
cor.plot + effect.plot + plot_annotation(tag_levels = 'a')
ggsave(here('figures', 'DA-analysis1.png'), width = 8, height = 4.5)
```


```{r}
sessioninfo::session_info()
```

