---
title: "Data Profile"
author: "Yixuan Yang"
output:
  html_document:
    toc: true
    toc_float: true
    self_contained: true
---

# Set up

```{r message=FALSE}
rm(list = ls())
set.seed(2024)
library(here)
library(phyloseq)
library(tidyverse)
theme_set(theme_bw())
```

# Read in data

Both pilot and following study data are loaded

```{r}
ps <- read_rds(here('data','following_study','ps.rds'))
ps.pilot <- read_rds(here('data','pilot_study','ps.rds'))
sam <- data.frame(sample_data(ps))
```

# Data Profile

Check frequency of breed group between health condition and perform chi-square and proportion test

```{r}
tb <- sam %>% 
  filter(!is.na(Breed.Group..1.), is.na(Breed.Group..2.)) %>%
  xtabs(~Breed.Group..1. + Epileptic.or.Control, data = .)
tb
chisq.test(tb)
prop.test(tb)
```


Check frequency of sex between health condition and perform chi-square and proportion test

```{r}
tb <- xtabs(~Sex + Epileptic.or.Control, data = sam)
tb
chisq.test(tb)
```

Check frequency of Phenobarbital Administration between health condition and perform chi-square and proportion test

```{r}
tb <- xtabs(~Pheno.Y.N + Epileptic.or.Control, data = sam)
tb
```

Does Phenobarbital Administration effective?

```{r}
tb <- sam %>% 
  filter(Epileptic.or.Control == 'Epileptic') %>% 
  xtabs(~Pheno.Y.N + Seizure.Freedom..Y.N., data = .)
tb
chisq.test(tb)
```

Get distribution of Dogs age in month

```{r}
sam %>% 
  group_by(Epileptic.or.Control) %>% 
  summarise(mean= mean(Age..months.), `standard deviation` = sd(Age..months.))
```

# 16S rRNA Sequencing Profile

Double-check if microbial pattern is consistent with health condition. Here we draw the abundance plot by Phylum

```{r}
ps.all <- merge_phyloseq(ps, ps.pilot) %>% 
  tax_glom('Phylum') %>% 
  transform_sample_counts(function(x) x / sum(x)) %>% # transform to proportional
  psmelt()
ps.all$Phylum <- as.character(ps.all$Phylum)
ps.all$Phylum[ps.all$Abundance < 0.01] <- 'Others'
ps.all$study <- factor(ps.all$study, levels = c('pilot', 'present'))
```


```{r}
ggplot(data = ps.all) + 
  geom_col(aes(Sample, Abundance, fill = Phylum),position = "fill", width = 1) +
  theme(axis.text.x=element_blank(),
        axis.ticks.x=element_blank())  +
  scale_y_continuous(labels = scales::percent) +
  facet_grid(. ~ study, scales="free_x", space="free_x") +
  labs(y = 'Relative Abundance', x = 'Sample', fill = 'Phylum')
ggsave(here('figures','abundance_phylum.png'), width = 6, height = 4)
```

```{r}
phylum.composition <- ps.all %>% 
  split(.$study) %>%
  lapply(function(x) x %>% 
           group_by(Sample,Phylum) %>%
           summarise(Abundance = sum(Abundance)) %>%
           arrange(desc(Abundance)) %>% 
           pivot_wider(names_from = Phylum, values_from = Abundance, values_fill = 0))
phylum.composition %>% head()
```

```{r}
phylum.composition %>% lapply(summary)
```

```{r}
genus.ps <- ps %>% subset_taxa(Phylum == 'Actinobacteriota') %>% 
  transform_sample_counts(function(x) x / sum(x)) %>% 
  psmelt()
genus.ps$Genus[genus.ps$Abundance < 0.01] <- 'Others'
genus.ps %>% 
  group_by(Sample, Genus) %>% 
  summarise(Abundance = sum(Abundance), .groups = 'drop') %>% 
  group_by(Genus) %>% 
  summarise(mean = sum(Abundance)/98, `standard deviation` = sd(Abundance)) %>% 
  arrange(desc(mean))
```

```{r}
genus.ps <- ps %>% subset_taxa(Phylum == 'Firmicutes') %>% 
  transform_sample_counts(function(x) x / sum(x)) %>% 
  psmelt()
genus.ps$Genus[genus.ps$Abundance < 0.01] <- 'Others'
genus.ps %>% 
  group_by(Sample, Genus) %>% 
  summarise(Abundance = sum(Abundance), .groups = 'drop') %>% 
  group_by(Genus) %>% 
  summarise(mean = sum(Abundance)/98, `standard deviation` = sd(Abundance)) %>% 
  arrange(desc(mean))
```

```{r}
sessioninfo::session_info()
```




